\emph{dfoliatR}: An R package for detection and analysis of insect defoliation signals in tree rings

### 2020-03-05 paper_elsevier/paper_elsevier.Rmd
working on the tree overview

The `dfoliatR` library is unique among a growing suite `R` packages designed for dendrochronology. Stemming from the `dplR` library [@Bunn2008] that provides `R` the ability to read and write an array of tree-ring data formats, standaridize ring widths, build and evaluate chronologies, perform qaulity control (to name a few), one can now also measure, perform and check crossdating, and perform many analytical tests [@Bunn2010; @Jevsenak2018; @Lara2015; @Zang2015]. Tools for assessing stand dynamics and disturbance analyses are under rapid development, with new packages for assessing growth and release events [`TRADER`: @Altman2014], metrics of growth resilience [`pointRes`: @VanderMaaten-Theunissen2015], and fire history [`burnr`: @Malevich2018]. The key objective of `dfoliatR` is to provide tools to identify and analyse insect defoliation and outbreak events by building on the methods employed the FORTRAN program OUTBREAK [@Swetnam1989]. What sets `dfoliatR` apart from packages such as `TRADER` is that it explicitly performs an indexing procedue on host-tree series to remove climatic and other non-defoliation related signals represented by non-host tree species. Insect defoliation signals are identified in the disturbance index by the duration and magnitude of negative departures.

`dfoliatR` draws upon data formats in `dplR` that are commonly employed by other tree-ring libraries. It uses and outputs data formats that faciliate the use packages embodied by `tidyverse` [@Wickham2019] that include efficient manipulation of data [`dplyr`: @Wickham2020dplyr] and graphics [`ggplot2`: @wickham2016ggplot2].

@@

Users can develop these data sets in software of their choosing, such as in `dplR` or ARSTAN. It is important that the host-tree data include only one tree-ring series per tree. Both `dplR` and ARSTAN have options for averaging mutliple sample series into a tree-level series. The tree-ring series and chronology can be read into `R` via several available `dplR` functions.

`dfoliatR` begins to identify defoliation events in individual trees by removing the climatic signal as represented by the non-host chronology from the host tree series. This indexing procedure creates a series -- the "growth suppression index" (GSI) -- in which disturbance is the predominant signal. The GSI is caluclated as 
\begin{align}
\textrm{GSI}_{i} = \textrm{H}_{i} - \left( \textrm{NH}_{i} - \overline{\textrm{NH}} \right) \frac{\sigma_{\textrm{H}}}{\sigma_{\textrm{NH}}} \
\end{align}
where H and NH are the host tree series and the non-host chronology, in year i, respectively [@Swetnam1985; @Swetnam1989]. Only years in which the individual host-tree series and the non-host chronology overlap are used in Equation 1. The non-host chronology is scaled by its mean ($\overline{\textrm{NH}} \approx 1.0$) and multiplied by the ratio of host and non-host standard deviations ($\frac{\sigma_{\textrm{H}}}{\sigma_{\textrm{NH}}}$), which approximates the variance of the host tree series. This "corrected" non-host chronology is subtracted from the host-tree series.

Negative departures in the normalized growth suppression index that surpass user-defined thresholds in duration and magnitude will be defined as *defoliation events*.

@@

Note that these methods of separating tree- vs site-level disturbance categories is a major departure from the OUTBREAK program. In OUTBREAK the two levels of analysis are combined and users have more limited control of thresholds to define defoliation events versus outbreaks.

@@

Usage
==========================

@@

##Site-level events

>>>>

The `dfoliatR` library adds to a growing suite `R` packages designed for dendrochronology. Stemming from the `dplR` library [@Bunn2008] that enables `R` users to read and write an array of tree-ring data formats, standaridize ring widths, build and evaluate chronologies, perform qaulity control (to name a few), one can now also measure ring widths [@Shi2019; @Lara2015], perform and check crossdating [@Bunn2010], and perform many analytical tests [@Jevsenak2018; @Zang2015]. Tools for assessing stand dynamics and disturbance analyses are under rapid development, with new packages for assessing growth and release events [`TRADER`: @Altman2014], metrics of growth resilience [`pointRes`: @VanderMaaten-Theunissen2015], and fire history [`burnr`: @Malevich2018]. The key objective of `dfoliatR` is to provide tools to identify and analyse insect defoliation and outbreak events by building on the methods employed the FORTRAN program OUTBREAK [@Swetnam1989]. What sets `dfoliatR` apart from packages such as `TRADER` is that it explicitly performs an indexing procedue on host-tree series to remove climatic and other non-defoliation related signals represented by non-host tree species. Insect defoliation signals are identified in the disturbance index by the duration and magnitude of negative departures.

`dfoliatR` draws upon data formats in `dplR` that are commonly employed by other tree-ring libraries. It uses and outputs data formats that faciliate the use packages embodied by `tidyverse` [@Wickham2019] that include efficient data manipulation [`dplyr`: @Wickham2020dplyr] and graphics [`ggplot2`: @wickham2016ggplot2].

@@

Users can develop these data sets in software of their choosing, such as `dplR` or ARSTAN. It is important that the host-tree data include only one tree-ring series per tree. Both `dplR` and ARSTAN have options for averaging mutliple sample series into a tree-level series. The tree-ring series and chronology can be read into `R` via several available `dplR` functions.

At the heart of `dfoliatR` lies two functions: `defoliate_trees()` and `outbreak()`. These identify defoliation event son individual trees and thenc omposite across mutliple trees for a given site to identify outbreak events. It is up to the researcher to distinguish these separate spatial scales of analysis, `dfoliatR` then provides a set of functions to visualize and statistically summarize tree-level defoliation and site-level outbreak events, which will be discussed further.

## Identifying Defoliation of Trees

The `defoliate_trees()` function will often be the point of entry to the `dfoliatR` library. It performs two processes: removing the non-defoliatation growth signal from the host-tree series and identifying defoliation events for each host tree. The climatic or non-defoliation signals in each host-tree series is represented by the non-host chronology or a climate reconstruction. This indexing procedure generates the "growth suppression index" (GSI) in which disturbance is the predominant signal. The GSI is calculated as in @Swetnam1985 and @Swetnam1989, 
\begin{align}
\textrm{GSI}_{i} = \textrm{H}_{i} - \left( \textrm{NH}_{i} - \overline{\textrm{NH}} \right) \frac{\sigma_{\textrm{H}}}{\sigma_{\textrm{NH}}} \
\end{align}
where H and NH are the host-tree series and the non-host chronology, in year i, respectively. Only the common period between each host-tree series and the non-host chronology are used in Equation 1. The non-host chronology is scaled by its mean ($\overline{\textrm{NH}} \approx 1.0$) and multiplied by the ratio of host and non-host standard deviations ($\frac{\sigma_{\textrm{H}}}{\sigma_{\textrm{NH}}}$), which approximates the variance of the host tree series. This "corrected" non-host chronology is subtracted from the host-tree series.

Negative departures in the normalized GSI that surpass user-defined thresholds in duration and magnitude are defined as *defoliation events*. As in OUTBREAK, defoliation events have a single year of maximum departure (the variable 'max_reduction' in `defoliate_trees()`) that is used to assess the magnitude of growth suppression. The default setting is -1.28 (in units of standard deviation), which was previously determined as representative of WSBW effects (**cite**). Duration is assessed by examining sequences of negative GSI before and after the year of maximum departure. Each defoliation event is allowed one positive excersion on each side of the maximum departure year. Duration is assessed across the entire sequence that includes up to two positive excursions. The default duration is eight years, following OUTBREAK, as is commonly used in WSBW studies (**cite**). Defoliator insects vary in their durations of defoliations and outbreaks, so researchers should exmaine their data and perform sensitivity tests to to find a representative event duration.

Researchers are given the options to extend defoliation events by bridging between successive events and allowing potentially short-duration events that occur at the end of the series. Bridging allows multiple events that are separated by a single year to be combined and considered as one defoliation event. We urge researchers to use this option carefully, as there is no setting to limit the number of potentially bridged events. Series end events is an option that can be used when the trees were sampled during a known defoliation event, which in many cases is what led the to the site being sampled for the reconstruction. This option eliminates the duration parameter, but all other thresholds, and bridging, still apply. These options are not available in OUTBREAK.

## Identifying Outbreak Events

@@

@@

Example Usage
==========================

@@

## Site-level events

### 2020-03-05 paper_elsevier/paper_elsevier.tex
working on the tree overview

The \texttt{dfoliatR} library is unique among a growing suite \texttt{R}
packages designed for dendrochronology. Stemming from the \texttt{dplR}
library (Bunn 2008) that provides \texttt{R} the ability to read and
write an array of tree-ring data formats, standaridize ring widths,
build and evaluate chronologies, perform qaulity control (to name a
few), one can now also measure, perform and check crossdating, and
perform many analytical tests (Bunn 2010, Lara et al. 2015, Zang and
Biondi 2015, Jevšenak and Levanič 2018). Tools for assessing stand
dynamics and disturbance analyses are under rapid development, with new
packages for assessing growth and release events (\texttt{TRADER}:
Altman et al. 2014), metrics of growth resilience (\texttt{pointRes}:
Maaten-Theunissen et al. 2015), and fire history (\texttt{burnr}:
Malevich et al. 2018). The key objective of \texttt{dfoliatR} is to
provide tools to identify and analyse insect defoliation and outbreak
events by building on the methods employed the FORTRAN program OUTBREAK
(Swetnam and Lynch 1989). What sets \texttt{dfoliatR} apart from
packages such as \texttt{TRADER} is that it explicitly performs an
indexing procedue on host-tree series to remove climatic and other
non-defoliation related signals represented by non-host tree species.
Insect defoliation signals are identified in the disturbance index by
the duration and magnitude of negative departures.

\texttt{dfoliatR} draws upon data formats in \texttt{dplR} that are
commonly employed by other tree-ring libraries. It uses and outputs a
data formats that faciliate the use packages embodied by
\texttt{tidyverse} (Wickham et al. 2019) that include efficient data
wrangling (\texttt{dplyr}: Wickham et al. 2020) and graphics
(\texttt{ggplot2}: Wickham 2016).

@@

Users can develop these data sets in software of their choosing, such as
in \texttt{dplR} or ARSTAN. It is important that the host-tree data
include only one tree-ring series per tree. Both \texttt{dplR} and
ARSTAN have options for averaging mutliple sample series into a
tree-level series. The tree-ring series and chronology can be read into
\texttt{R} via several available \texttt{dplR} functions.

\texttt{dfoliatR} begins to identify defoliation events in individual
trees by removing the climatic signal as represented by the non-host
chronology from the host tree series. This indexing procedure creates a
series -- the ``growth suppression index'' (GSI) -- in which disturbance
is the predominant signal. The GSI is caluclated as \begin{align}
\textrm{GSI}_{i} = \textrm{H}_{i} - \left( \textrm{NH}_{i} - \overline{\textrm{NH}} \right) \frac{\sigma_{\textrm{H}}}{\sigma_{\textrm{NH}}} \
\end{align} where H and NH are the host tree series and the non-host
chronology, in year i, respectively (Swetnam et al. 1985, Swetnam and
Lynch 1989). Only years in which the individual host-tree series and the
non-host chronology overlap are used in Equation 1. The non-host
chronology is scaled by its mean
(\(\overline{\textrm{NH}} \approx 1.0\)) and multiplied by the ratio of
host and non-host standard deviations
(\(\frac{\sigma_{\textrm{H}}}{\sigma_{\textrm{NH}}}\)), which
approximates the variance of the host tree series. This ``corrected''
non-host chronology is subtracted from the host-tree series.

Negative departures in the normalized growth suppression index that
surpass user-defined thresholds in duration and magnitude will be
defined as \emph{defoliation events}.

@@

\hypertarget{usage}{%
\section{Usage}\label{usage}}

@@

\#\#Site-level events

@@

>>>>

The \texttt{dfoliatR} library adds to a growing suite \texttt{R}
packages designed for dendrochronology. Stemming from the \texttt{dplR}
library (Bunn 2008) that enables \texttt{R} users to read and write an
array of tree-ring data formats, standaridize ring widths, build and
evaluate chronologies, perform qaulity control (to name a few), one can
now also measure ring widths (Lara et al. 2015, Shi and Xiang 2019),
perform and check crossdating (Bunn 2010), and perform many analytical
tests (Zang and Biondi 2015, Jevšenak and Levanič 2018). Tools for
assessing stand dynamics and disturbance analyses are under rapid
development, with new packages for assessing growth and release events
(\texttt{TRADER}: Altman et al. 2014), metrics of growth resilience
(\texttt{pointRes}: Maaten-Theunissen et al. 2015), and fire history
(\texttt{burnr}: Malevich et al. 2018). The key objective of
\texttt{dfoliatR} is to provide tools to identify and analyse insect
defoliation and outbreak events by building on the methods employed the
FORTRAN program OUTBREAK (Swetnam and Lynch 1989). What sets
\texttt{dfoliatR} apart from packages such as \texttt{TRADER} is that it
explicitly performs an indexing procedue on host-tree series to remove
climatic and other non-defoliation related signals represented by
non-host tree species. Insect defoliation signals are identified in the
disturbance index by the duration and magnitude of negative departures.

\texttt{dfoliatR} draws upon data formats in \texttt{dplR} that are
commonly employed by other tree-ring libraries. It uses and outputs data
formats that faciliate the use packages embodied by \texttt{tidyverse}
(Wickham et al. 2019) that include efficient data manipulation
(\texttt{dplyr}: Wickham et al. 2020) and graphics (\texttt{ggplot2}:
Wickham 2016).

@@

Users can develop these data sets in software of their choosing, such as
\texttt{dplR} or ARSTAN. It is important that the host-tree data include
only one tree-ring series per tree. Both \texttt{dplR} and ARSTAN have
options for averaging mutliple sample series into a tree-level series.
The tree-ring series and chronology can be read into \texttt{R} via
several available \texttt{dplR} functions.

At the hear of \texttt{dfoliatR} lies two functions:
\texttt{defoliate\_trees()} and \texttt{outbreak()}. These identify
defoliation event son individual trees and thenc omposite across
mutliple trees for a given site to identify outbreak events. It is up to
the researcher to distinguish these separate spatial scales of analysis,
\texttt{dfoliatR} then provides a set of functions to visualize and
statistically summarize tree-level defoliation and site-level outbreak
events, which will be discussed further.

\hypertarget{identifying-defoliation-of-trees}{%
\subsection{Identifying Defoliation of
Trees}\label{identifying-defoliation-of-trees}}

The \texttt{defoliate\_trees()} function will usually be used first. It
performs two processes: removing the non-defoliatation growth signal
from the host-tree series and identifying defoliation events for each
host tree. The climatic or non-defoliation signals in each host-tree
series is represented by the non-host chronology or a climate
reconstruction. This indexing procedure generates the ``growth
suppression index'' (GSI) in which disturbance is the predominant
signal. The GSI is calculated as in Swetnam et al. (1985) and Swetnam
and Lynch (1989), \begin{align}
\textrm{GSI}_{i} = \textrm{H}_{i} - \left( \textrm{NH}_{i} - \overline{\textrm{NH}} \right) \frac{\sigma_{\textrm{H}}}{\sigma_{\textrm{NH}}} \
\end{align} where H and NH are the host tree series and the non-host
chronology, in year i, respectively. Only the common period between each
host-tree series and the non-host chronology are used in Equation 1. The
non-host chronology is scaled by its mean
(\(\overline{\textrm{NH}} \approx 1.0\)) and multiplied by the ratio of
host and non-host standard deviations
(\(\frac{\sigma_{\textrm{H}}}{\sigma_{\textrm{NH}}}\)), which
approximates the variance of the host tree series. This ``corrected''
non-host chronology is subtracted from the host-tree series.

Negative departures in the normalized GSI that surpass user-defined
thresholds in duration and magnitude are defined as \emph{defoliation
events}. As in OUTBREAK, defoliation events have a single year of
maximum departure (the variable `max\_reduction' in
\texttt{defoliate\_trees()}) that is used to assess the magnitude of
growth suppression. The default setting is -1.28 (in units of standard
deviation), which was previously determined as representative of WSBW
effects (\textbf{cite}). Duration is assessed by examining sequences of
negative GSI before and after the year of maximum departure. Each
defoliation event is allowed one positive excersion on each side of the
maximum departure year. Duration is assessed across the entire sequence
that includes up to two positive excursions. The default duration is
eight years, following OUTBREAK, as is commonly used in WSBW studies
(\textbf{cite}). Defoliator insects vary in their durations of
defoliations and outbreaks, so researchers should exmaine their data and
perform sensitivity tests to to find a representative event duration.

Researchers are given the options to extend defoliation events by
bridging between successive events and allowing potentially
short-duration events that occur at the end of the series. Bridging
allows multiple events that are separated by a single year to be
combined and considered as one defolaition event. Each may have up to
five years with positive growth (\textbf{CHECK}). Series end events is
an option that can be used when the trees were sampled during a known
defoliation event, which in many cases is what led the to the site being
sampled for the reconstruction. This option eliminates the duration
parameter, but all other thresholds, and bridging, still apply. These
options are not available in OUTBREAK.

\hypertarget{identifying-outbreak-events}{%
\subsection{Identifying Outbreak
Events}\label{identifying-outbreak-events}}

@@

\hypertarget{example-usage}{%
\section{Example Usage}\label{example-usage}}

@@

\hypertarget{site-level-events}{%
\subsection{Site-level events}\label{site-level-events}}

@@

\leavevmode\hypertarget{ref-Shi2019}{}%
Shi, J., and W. Xiang. 2019. MtreeRing: A shiny application for
automatic measurements of tree-ring widths on digital images.
